Characterizing dynamical transitions by statistical complexity measures based on ordinal pattern transition networks

### 2021-02-10 manuscript_Huang_rev.tex
Reik's changes.

In addition to the proper normalization, we need to consider self-loops that may affect the numerical estimation of $\mathbf{W}$ especially for time series from continuous systems with a given sampling rate. For example, it has been demonstrated in Ref.~\cite{zhangSciRep2017} that there are about $99\%$ self-loops in time series of the R\"ossler system when integrated with a step size $h = 0.01$, while the about $1\%$ non-self-loops are hidden in $\mathbf{W}$. From this example, one may easily see that self-loops are related to the serial correlation of the time series under study \cite{BorgesAMC2019}. {\color{red} From the viewpoint of statistical measures of complex network theory}, neglecting self-loops in $\mathbf{W}$ emphasizes the transition behavior between different patterns, which has therefore been adopted in most existing studies of complex networks for computational simplicity and theoretical concerns \cite{CostaADPhy2007}. Accordingly, we also remove self-loops in the present study before computing $\mathbf{W}$ while acknowledging that in other cases of {\color{red} highly correlated} stochastic processes and discrete systems, self-loops have been included in the analysis showing some interesting results \cite{BorgesAMC2019}. Arguably, whether to remove or to consider self-loops depends on the particular process under study. In any of the applications discussed in the rest of this paper, our results do not change qualitatively when including self-loops{\color{red} (see Supplementary Information Fig.~S1 to S3 for some illustrative examples).}

@@

The situation becomes even more challenging when our modified SCMs are considered, since they are based on transition frequencies instead of pattern frequencies. For $D!$ different ordinal patterns for embedding dimension $D$, there may exist $D!^2$ possible source--destination pairs. However, if we consider $\tau=1$ and transitions corresponding to shifting the time index by one time step only, it is clear that from the $D$ components of the source permutation, $D-1$ also contribute to the corresponding destination pattern without changing their mutual order relationships. Hence, the new state contributing to the destination pattern can only lead to at most $D$ different order patterns, i.e., we have at most $D\cdot D!$ different pattern transitions. However, for other cases like $\tau>1$ and/or larger time shifts between ordinal patterns to be considered, up to $D!^2$ different pairs of patterns and, hence, transitions could occur in the worst case. In general, we conclude that the far higher number of pattern transitions as compared to patterns themselves puts severe restrictions to the choice of $D$ in practical applications of the new SCMs in the context of time series of a given length $L$.
}

@@

\subsection{\textcolor{red}{Effect} of embedding parameters} \label{sec:embeddings}
There are two important algorithmic parameters in constructing OPTNs -- the embedding parameters $D$ and $\tau$. The choice of $D$ and $\tau$ can markedly affect the results for the associated SCMs since $D$ and $\tau$ should compromise the essential time scales of the dynamics in the underlying time series. In common applications of ordinal pattern analysis (e.g., the estimation of the permutation entropy $\mathcal{S}_O$), the choice of optimal (or just pragmatic) values of $D$ and $\tau$ depends on the field of application as systematically summarized in Ref.~\cite{Riedl2013}. {\color{red}Throughout the remainder of this work, we will consider $D$ as a free parameter, which is varied in the range of $D \in \{2,\ldots,7\}$, although for the sake of brevity, we will mainly show the results for $D = 6$ as illustration. The associated choice of the embedding delay $\tau$ is commonly less crucial in nonlinear time series analysis methods. In our numerical examples discussed below, the corresponding selection will be mainly guided by the attempt to highlight the fact that lower entropy values are expected for periodic as compared to chaotic behavior. In order to cope with a single parameter only, we will particularly use $\tau=D$ for the artificial (deterministic and stochastic) model systems considered in the following, independent of the particular value of $D$. We emphasize that this choice is not based on specific considerations of performance optimization, but simply reflects convenience and serves solely illustrative purposes.}

@@

As an illustrative example highlighting the important roles of $D$ and $\tau$ for ordinal pattern based \textcolor{red}{time series} analysis, we consider a period-4 solution of the logistic map 
\begin{equation}
x_{i+1}=rx_i(1-x_i) 
\end{equation}
\noindent
with $r= 3.5$. {\color{red}In this example, the choice of $\tau = D$ reflects the particular role of specific choices of the embedding parameters in the context of low-order periodic windows (here, period-4 and period-3) in the bifurcation sequence of the logistic map. As expected,} Fig.~\ref{fig:embed} demonstrates that the resulting ordinal pattern sequences are significantly affected by the different parameter settings, which further result in different permutation entropy values $\mathcal{H}_O$. For instance, only one unique ordinal pattern is observed when $\tau = D = 4$ (Fig.~\ref{fig:embed}(c, g)) resulting in $\mathcal{H}_O = 0$, which is due to the coverage of exactly one full period by the particular embedding parameters. Other values of $D$ and $\tau$ however yield non-zero entropy values. The results of Fig.~\ref{fig:embed} may raise concern since different choices of $D$ and $\tau$ will change the placement of the system in the complexity--entropy plane. %{\color{red}, which is an open problem for ordinal pattern analysis \cite{AmigoPTRSA2014,Riedl2013}.} 
In the special case of the logistic map, we do not have a unique choice of $D$ and $\tau$ for {\color{red}all different} periodic windows when the control parameter $r$ is varied. When the control parameter $r$ is changed, we should \textcolor{red}{rather} focus on the transitions between different dynamical regimes. 
% In the following, we will show that the results are robust when the control parameter is systematically changed while keeping the same embedding parameters.

@@

As a first step, we consider four representative regimes that can be observed at particular values of the control parameter $r$, corresponding to either periodic dynamics or chaotic dynamics close to band merging, in some laminar state, and at the outer crisis. In all examples, we will consistently use the embedding parameters $D = \tau = 6$. 
%In general, we emphasize that a higher value of $D$ leads to a larger variety of ordinal patterns and, hence, a more robust statistics. On the other hand, for evaluating the statistical properties of either pattern frequencies or transition frequencies, this comes on the cost of requiring increasingly longer time series. 
In the following, we will use realizations of the logistic map comprising $L = 10^6$ iterations (after removing the first $1000$ iterations of each realization which might reflect transient behavior of the system before reaching its respective attractor)\textcolor{red}{, which meets the conditions for the necessary time series length for $D=6$ as discussed in Sect.~\ref{sec:tslength}}.

@@

In the context of finite sample size effects, we note that for the particular example of the logistic map, one could easily generate realizations of much longer lengths. Our numerical results show, however, that longer realizations of a length of up to $L\leq5\cdot 10^6$ do not change the obtained entropy and SCM \emph{estimates} significantly (not shown). From a practical perspective, two aspects call for attention in interpreting the obtained entropy and SCM values quantitatively. On the one hand, due to the finite numerical precision of the computed trajectories, any chaotic regime may in fact turn out to appear as periodic (with arbitrarily long periods); this can lead to numerical artifacts that will be further discussed in Sec.~\ref{sec:transi}. On the other hand, there may exist asymptotic analytical results for the case of $L\to\infty$ and $D\to\infty$, e.g., for the case of $r=4$ where one would expect $\mathcal{H}_O\to 1$ and $\mathcal{C}_O\to 0$, which substantially differ from our numerical results. In this regard, all estimated values of the different entropy and complexity measures always need to be considered as finite sample estimates that may well deviate from their expected asymptotic values and thereby depend on the specific values of $L$ and $D$ available for the analysis. This is, however, a common feature of other existing symbolic dynamics based entropies and SCMs.

@@

In previous works, SCMs have been applied to distinguish chaos from noise by employing the concept of complexity--entropy plane~\cite{rossoPRL2007}. Generalizing this idea to our OPTN based SCMs, we can define three such planes showing the statistical complexity measures ($\mathcal{C}_O$, $\mathcal{C}_W$ and $\mathcal{C}_E$) as a function of the corresponding normalized entropy values ($\mathcal{H}_O$, $\mathcal{H}_W$ and $\mathcal{H}_E$). The three SCMs characterize both randomness and correlation structure in a time series, which as a consequence result in a non-trivial dependence on the associated entropy values. Specifically, chaotic systems present high complexity while stochastic systems have lower values of complexity, hence ideally appearing in distinct regions of the complexity--entropy plane~\cite{rossoPRL2007}. Furthermore, {\color{red}it has been shown previously that at a given entropy value,} the range of possible SCM values is bound by a minimum $\mathcal{C}_{min}$ and a maximum $\mathcal{C}_{max}$. A general algorithm for computing these bounds has been provided in Ref.~\cite{martinPhyA2006}.

Figure~\ref{fig:CElogistic} shows the complexity--entropy planes for embedding parameters $D = \tau$ varied between $4$ and $6$. We emphasize that the results are qualitatively similar \textcolor{red}{for $D=\tau=3,\ 7$ as well (not shown)}. Furthermore, it is notable that the values of $\mathcal{C}_{max}$ depend on the embedding dimension $D$ since $D$ determines the number of patterns and pattern transitions which are considered in the definition of SCMs. {\color{red}The curves of $\mathcal{C}_{max}$ for different embeddings have been further annotated in Fig. \ref{fig:CElogistic}. In contrast, the $\mathcal{C}_{min}$ values are close to each other, showing less dependence on $D$, and therefore they are not annotated. }
\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{CompEntropy_LogisticHC.pdf}
\caption{\small{(Color online) Three different complexity--entropy planes (based on the Jensen-Shannon divergence) illustrating the behavior of the three SCMs for example series of the logistic map with varying control parameter $r$: (a, d) $\mathcal{C}_O$ versus $\mathcal{H}_O$, (b, e) $\mathcal{C}_{W}$ versus $\mathcal{H}_{W}$, and (c, f) $\mathcal{C}_{E}$ versus $\mathcal{H}_{E}$. The {\color{red} dotted lines correspond to $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$ in dependence on the respective entropy for $D=4$, and the dashed lines are $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$ for $D=5$. Panels (a-c) show the results for embedding dimensions $D = 4,5$ with annotations for the respective $\mathcal{C}_{max}$ only} and (d-f) for $D = 6$. In panels (a, d), the bottom-right rectangles of dotted red lines indicate the regions for the traditional pattern frequency based SCM analysis where the differentiation between stochastic and deterministic--chaotic dynamics is ambiguous. However, the lower left rectangles of (b, c, e, f) correspond to regions of lower complexities. }  \label{fig:CElogistic}}
\end{figure*}

First of all, we recover the known result that pattern frequency based SCM values lie close to the maximum $\mathcal{C}_{max}$, while the displacements of the estimated values from their theoretical upper bounds depend on the embedding dimension $D$ as shown in Fig.~\ref{fig:CElogistic}(a,d). Most notably, we find many combinations of complexity--entropy values in the lower right part of the plane (high entropy, low complexity), which have been highlighted by rectangles of dotted lines. This region of the plane is challenging for SCMs to distinguish between chaotic time series and random noise~\cite{BorgesAMC2019} and is in our case occupied by values corresponding to chaotic regimes of the logistic map at large $r$ values. The corresponding ambiguity results from the fact that both $\mathcal{H}_O$ and $\mathcal{C}_O$ only take static information in the sense that only pattern frequencies have been considered in their definitions.

For our new pattern transition based SCMs, we find that in the complexity--entropy planes $\mathcal{H}_W \times \mathcal{C}_W$ and $\mathcal{H}_E \times \mathcal{C}_E$, the results for the logistic map again closely align with the theoretical upper bounds $\mathcal{C}_{max}$. When increasing the embedding dimension, the estimated values approach the theoretical $\mathcal{C}_{max}$ curve more and more closely and essentially cover exclusively the left uprising branch of the curve for the SCM based solely on transition frequencies (Figs.~\ref{fig:CElogistic}(b, e) and \ref{fig:CElogistic}(c, f)), which distinctively differs from the original permutation entropy based SCM. Specifically, the definition of $\mathcal{C}_{E}$ and $\mathcal{H}_{E}$ takes information on both, pattern frequencies and transition frequencies into account, which results in the obtained SCM values aligning even more closely with $\mathcal{C}_{max}$ already at lower embedding dimensions (Fig.~\ref{fig:CElogistic}(c, f)). In our opinion, the most interesting result is, however, that we do not find any values in the lower right part of the plane, which for the classical SCM based on permutation entropy would present the challenging region of possible ambiguity between deterministic-chaotic and stochastic dynamics.

{\color{red}On the other hand, there are many values in the lower left corner of the plane showing small values of $\mathcal{C}_{W}$ and $\mathcal{C}_{E}$ (Fig. \ref{fig:CElogistic}(b, c, e, f)). This region corresponds to regular orbits of low complexities. Along with the choice of embedding parameters, the length of time series and long term accumulations of the computation precisions play important roles in the lower left region of the complexity planes, which will be further further discussed in Sec. \ref{sec:transi}. Out of the left and right corners, complexities of $\mathcal{C}_{W}$ and $\mathcal{C}_{E}$ show consistent trends along with the degree of chaoticity as will be explained later in Sec. \ref{sec:transi}.} In general, we therefore suggest that it is the deterministic transition behavior among patterns that allows an improved distinction from random noise.

{\color{red}In addition, we observe a linear correlation between $\mathcal{C}_W$ and $\mathcal{H}_{W}$ (Fig. \ref{fig:CElogistic}(e)), respectively, $\mathcal{C}_{E}$ and $\mathcal{H}_{E}$ (Fig. \ref{fig:CElogistic}(f)). However, we note that this linear correlation appears only for large embedding $D = 6$ and coincides with the up-rising brach of the maximal complexity curve $\mathcal{C}_{max}$, which is expected by the theory \cite{SMCbook2010}.}

{\color{red} The discriminatory power of different SCMs have been further compared in Fig. \ref{fig:CHlogCmax}, which characterize how well the alignments are to the maximal complexity $\mathcal{C}_{max}$. In particular, for a given $\mathcal{H}$ we compute a normalized complexity $\mathcal{C}^{N} = (\mathcal{C} - \mathcal{C}_{min}) / (\mathcal{C}_{max} - \mathcal{C}_{min})$. In consequence, $\mathcal{C}^{N}$ is bounded in the interval of $[0, 1]$ and the alignments along $\mathcal{C}_{max}$ are reflected by large values of $\mathcal{C}^{N}$. The results are shown in Fig. \ref{fig:CHlogCmax}. For the traditional $\mathcal{C}_{O}$, again we find there are many points in the right half plane (Fig. \ref{fig:CHlogCmax}(a) for all embedding dimensions. In contrast,  there are no points in the region of $\mathcal{H} >0.8$ for $\mathcal{C}_W$ and $\mathcal{C}_E$ (Fig. \ref{fig:CHlogCmax}(b,c)). Note that in Fig. \ref{fig:CHlogCmax}(c), the dots larger than 1 are numerical artefacts when the control parameters $r$ are close bifurcation points resulting in long transients (as will be explained by Fig. \ref{fig:transient} in Sec. \ref{sec:transi}).  
\begin{figure*}
        \centering
        \includegraphics[width=2\columnwidth]{CompEntropyCNormalized_Logistic.pdf}
\caption{(Color online) Normalized complexity -- entropy planes of three SCMs for the logistic map, which are normalized by $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$  at a given $\mathcal{H}$. (a) $\mathcal{C}_O$, (b) $\mathcal{C}_W$, and (c) $\mathcal{C}_E$. In (c), the dots larger than 1 are numerical artefacts when the control parameters $r$ are close bifurcation points resulting in long transients. Embedding dimensions are indicated by legends.  \label{fig:CHlogCmax}}
\end{figure*}
}

\subsubsection{Fractional Brownian motion and fractional Gaussian noise}
In order to further analyze the capability of our modified complexity--entropy planes to distinguish chaotic from stochastic dynamics, we compare the results for the logistic map with those obtained for fractional Brownian motion (fBm), i.e., stochastic processes with long-range temporal correlations. Specifically, for an fBm process the long-range correlations of the process are uniquely characterized by the Hurst exponent $H$ when positively correlated (persistence) for $1/2 < H < 1$, while being suppressed (anti-persistence) for $0< H < 1/2$. $H=1/2$ corresponds to the classical Brownian motion. Similar to the previous case of the logistic map with varying control parameter $r$, we generate time series for varying $H \in (0, 1)$ with a step size of $\Delta H=0.05$ and present the resulting complexity--entropy planes in Fig.~\ref{fig:CEfBm}. Moreover, since fBm series in the persistent regime are nonstationary, we further transform all fBm realizations into stationary time series by employing first-order difference filtering, i.e., by considering the increments $x_{i+1}- x_i$. The transformed series are commonly referred to as fractional Gaussian noise (fGn), and the associated complexity--entropy planes are shown in Fig.~\ref{fig:CEfGn}. Notably, fGn retains the long-range correlations and Gaussian probability density function (PDF) of the underlying fBm process. For a discussion on practical concerns regarding time delay embedding of fBm and fGn processes, we refer to Ref.~\cite{Zou2015}. 
\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{CompEntropy_fBm.pdf}
\caption{\small{(Color online) Same as in Fig.~\ref{fig:CElogistic}, but for fractional Brownian motion with a Hurst exponent $H \in (0, 1)$ varying with a step size of $\Delta H=0.05$. Note that the normalized entropy is generally decreasing with increasing $H$ (red arrow).}  \label{fig:CEfBm}}
\end{figure*}

>>>>

In addition to the proper normalization, we need to consider self-loops that may affect the numerical estimation of $\mathbf{W}$ especially for time series from continuous systems with a given sampling rate. For example, it has been demonstrated in Ref.~\cite{zhangSciRep2017} that there are about $99\%$ self-loops in time series of the R\"ossler system when integrated with a step size $h = 0.01$, while the about $1\%$ non-self-loops are hidden in $\mathbf{W}$. From this example, one may easily see that self-loops are related to the serial correlation of the time series under study \cite{BorgesAMC2019}. {\color{red} From the viewpoint of statistical measures of complex network theory}, neglecting self-loops in $\mathbf{W}$ emphasizes the transition behavior between different patterns, which has therefore been adopted in most existing studies of complex networks for computational simplicity and theoretical concerns \cite{CostaADPhy2007}. Accordingly, we also remove self-loops in the present study before computing $\mathbf{W}$ while acknowledging that in other cases of {\color{red} highly correlated} stochastic processes and discrete systems, self-loops have been included in the analysis showing some interesting results \cite{BorgesAMC2019}. Arguably, whether to remove or to consider self-loops depends on the particular process under study. In any of the applications discussed in the rest of this paper, our results do not change qualitatively when including self-loops {\color{red}(see Supplementary Material Sec.~SM1 with Figs.~S1 to S3 for some illustrative examples).}

@@

The situation becomes even more challenging when our modified SCMs are considered, since they are based on transition frequencies instead of pattern frequencies. For $D!$ different ordinal patterns for embedding dimension $D$, there may exist $D!^2$ possible source--destination pairs. However, if we consider $\tau=1$ and transitions corresponding to shifting the time index by one time step only, it is clear that from the $D$ components of the source permutation, $D-1$ also contribute to the corresponding destination pattern without changing their mutual order relationships. Hence, the new state contributing to the destination pattern can only lead to at most $D$ different ordinal patterns, i.e., we have at most $D\cdot D!$ different pattern transitions. However, for other cases like $\tau>1$ and/or larger time shifts between ordinal patterns to be considered, up to $D!^2$ different pairs of patterns and, hence, transitions could occur in the worst case. In general, we conclude that the far higher number of pattern transitions as compared to patterns themselves puts severe restrictions to the choice of $D$ in practical applications of the new SCMs in the context of time series of a given length $L$.
}

@@

\subsection{\textcolor{red}{Effect} of embedding parameters} \label{sec:embeddings}
There are two important algorithmic parameters in constructing OPTNs -- the embedding parameters $D$ and $\tau$. The choice of $D$ and $\tau$ can markedly affect the results for the associated SCMs since $D$ and $\tau$ should compromise the essential time scales of the dynamics in the underlying time series. In common applications of ordinal pattern analysis (e.g., the estimation of the permutation entropy $\mathcal{S}_O$), the choice of optimal (or just pragmatic) values of $D$ and $\tau$ depends on the field of application as systematically summarized in Ref.~\cite{Riedl2013}. {\color{red}Throughout the remainder of this work, we will consider $D$ as a free parameter, which is varied in the range of $D \in \{2,\ldots,7\}$, although for the sake of brevity, we will mainly show the results for $D = 6$ as illustration. The associated choice of the embedding delay $\tau$ is commonly less crucial in nonlinear time series analysis methods. In our numerical examples discussed below, the corresponding selection will be mainly guided by the attempt to highlight the fact that lower entropy values are expected for periodic as compared to chaotic behavior. In order to cope with a single parameter only, we will particularly use $\tau=D$ for the artificial (deterministic and stochastic) model systems considered in the following, independent of the particular value of $D$. We emphasize that this choice is not based on specific considerations of performance optimization, but simply reflects convenience and serves solely illustrative purposes. Two examples for results obtained with $D=6$ and $\tau=1$ for two well-known model systems can be found in the Supplementary Material Sec.~SM2, Figs.~S4 and S5.}

@@

As an illustrative example highlighting the important roles of $D$ and $\tau$ for ordinal pattern based \textcolor{red}{time series} analysis, we consider a period-4 solution of the logistic map 
\begin{equation}
x_{i+1}=rx_i(1-x_i) 
\end{equation}
\noindent
with $r= 3.5$. {\color{red}In this example, the choice of $\tau = D$ reflects the particular role of specific choices of the embedding parameters in the context of low-order periodic windows (here, period-4 and period-3) in the bifurcation sequence of the logistic map. As expected,} Fig.~\ref{fig:embed} demonstrates that the resulting ordinal pattern sequences are significantly affected by the different parameter settings, which further result in different permutation entropy values $\mathcal{H}_O$. For instance, only one unique ordinal pattern is observed when $\tau = D = 4$ (Fig.~\ref{fig:embed}(c, g)) resulting in $\mathcal{H}_O = 0$, which is due to the coverage of exactly one full period by the particular embedding parameters. Other values of $D$ and $\tau$ however yield non-zero entropy values. The results of Fig.~\ref{fig:embed} may raise concern since different choices of $D$ and $\tau$ will change the placement of the system in the complexity--entropy plane. In the special case of the logistic map, we do not have a unique choice of $D$ and $\tau$ for {\color{red}all different} periodic windows when the control parameter $r$ is varied. When the control parameter $r$ is changed, we should \textcolor{red}{rather} focus on the transitions between different dynamical regimes.

@@

As a first step, we consider four representative regimes that can be observed at particular values of the control parameter $r$, corresponding to either periodic dynamics or chaotic dynamics close to band merging, in some laminar state, and at the outer crisis. In all examples, we will consistently use the embedding parameters $D = \tau = 6$. In the following, we will use realizations of the logistic map comprising $L = 10^6$ iterations (after removing the first $1000$ iterations of each realization which might reflect transient behavior of the system before reaching its respective attractor)\textcolor{red}{, which meets the conditions for the necessary time series length for $D=6$ as discussed in Sect.~\ref{sec:tslength}}.

@@

In the context of finite sample size effects, we note that for the particular example of the logistic map, one could easily generate realizations of much longer lengths. Our numerical results show, however, that longer realizations of a length of up to $L\leq5\cdot 10^6$ do not change the obtained entropy and SCM \emph{estimates} significantly (not shown). From a practical perspective, two aspects call for attention in interpreting the obtained entropy and SCM values quantitatively. On the one hand, due to the finite numerical precision of the computed trajectories, any chaotic regime may in fact turn out to appear as periodic (with arbitrarily long periods); this can lead to numerical artifacts that will be further discussed in Sec.~\ref{sec:transi}. On the other hand, there may exist asymptotic analytical results for the case of $L\to\infty$ and $D\to\infty$, e.g., for the case of $r=4$ where one would expect $\mathcal{H}_O\to 1$ and $\mathcal{C}_O\to 0$, which substantially differ from our numerical results. In this regard, all estimated values of the different entropy and complexity measures always need to be considered as finite sample estimates that may well deviate from their expected asymptotic values and thereby depend on the specific values of $L$ and $D$ available for the analysis. This is, however, a feature that is also common to other existing symbolic dynamics based entropies and SCMs.

@@

In previous works, SCMs have been applied to distinguish chaos from noise by employing the concept of complexity--entropy plane~\cite{rossoPRL2007}. Generalizing this idea to our OPTN based SCMs, we can define three such planes showing the statistical complexity measures ($\mathcal{C}_O$, $\mathcal{C}_W$ and $\mathcal{C}_E$) as a function of the corresponding normalized entropy values ($\mathcal{H}_O$, $\mathcal{H}_W$ and $\mathcal{H}_E$). The three SCMs characterize both randomness and correlation structure in a time series, which as a consequence result\textcolor{red}{s} in a non-trivial dependence on the associated entropy values. Specifically, chaotic systems present high complexity while stochastic systems have lower values of complexity, hence ideally appearing in distinct regions of the complexity--entropy plane~\cite{rossoPRL2007}. Furthermore, {\color{red}it has been shown previously that at a given entropy value,} the range of possible SCM values is bound by a minimum $\mathcal{C}_{min}$ and a maximum $\mathcal{C}_{max}$. A general algorithm for computing these bounds has been provided in Ref.~\cite{martinPhyA2006}.

Figure~\ref{fig:CElogistic} shows the complexity--entropy planes for embedding parameters $D = \tau$ varied between $4$ and $6$. We emphasize that the results are qualitatively similar \textcolor{red}{for $D=\tau=3,\ 7$ as well (not shown)}. Furthermore, it is notable that the values of $\mathcal{C}_{max}$ depend on the embedding dimension $D$ since $D$ determines the number of patterns and pattern transitions which are considered in the definition of SCMs. {\color{red}Along with the specific results for the logistic map, Fig.~\ref{fig:CElogistic} also shows the theoretical limiting curves of $\mathcal{C}_{min}$ and $\mathcal{C}_{max}$ for three different embedding dimensions ($D=4,\ 5,\ 6$). While the lower bounds show only little variation with increasing $D$, the upper limits are more strongly affected and gradually shift upward for larger embedding dimensions.}

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{CompEntropy_LogisticHC.pdf}
\caption{\small{(Color online) Three different complexity--entropy planes (based on the Jensen-Shannon divergence) illustrating the behavior of the three SCMs for example series of the logistic map with varying control parameter $r$: (a, d) $\mathcal{C}_O$ versus $\mathcal{H}_O$, (b, e) $\mathcal{C}_{W}$ versus $\mathcal{H}_{W}$, and (c, f) $\mathcal{C}_{E}$ versus $\mathcal{H}_{E}$. The {\color{red} dotted lines correspond to $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$ in dependence on the respective entropy for $D=4$ (panels a-c), and the dashed lines indicate $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$ for $D=5$ (a-c) and $D=6$ (d-f), respectively.} In panels (a, d), the bottom-right rectangles of dotted red lines indicate the regions for the traditional pattern frequency based SCM analysis where the differentiation between stochastic and deterministic--chaotic dynamics is ambiguous. However, the lower left rectangles of (b, c, e, f) correspond to regions of lower complexities. }  \label{fig:CElogistic}}
\end{figure*}

\textcolor{red}{As a first general observation}, we recover the known result that \textcolor{red}{the established} pattern frequency based SCM values \textcolor{red}{for the logistic map} lie close to the \textcolor{red}{theoretical} maximum $\mathcal{C}_{max}$ \textcolor{red}{for a given entropy value} \textcolor{red}{(Fig.~\ref{fig:CElogistic}(a, d))}, while \textcolor{red}{some residual variation of the} displacement of the estimated values from \textcolor{red}{those} upper bounds \textcolor{red}{is retained for all considered values of} the embedding dimension $D$. Most notably, we find many combinations of complexity--entropy values in the lower \emph{right} part of the plane (high entropy, low complexity), which \textcolor{red}{is commonly indicative of stochastic processes and has been visually} highlighted by rectangles of dotted lines. This region of the plane is challenging for SCMs to distinguish between chaotic time series and random noise~\cite{BorgesAMC2019} and is in our case occupied by values corresponding to chaotic regimes of the logistic map at large $r$ values \textcolor{red}{(see also our previous comment on the expected asymptotic value for $r=4$)}. The corresponding ambiguity results from the fact that both $\mathcal{H}_O$ and $\mathcal{C}_O$ only take static information \textcolor{red}{into account} in the sense that only pattern frequencies have been considered in their definitions. \textcolor{red}{The observed behavior confirms previous findings that distinguishing (especially high-dimensional) chaotic from stochastic dynamics can be a challenging task in established ordinal pattern based time series analysis methods~\cite{AmigoPTRSA2014,Riedl2013}.}

For our new pattern transition based SCMs, we find that in the complexity--entropy planes $\mathcal{H}_W \times \mathcal{C}_W$ and $\mathcal{H}_E \times \mathcal{C}_E$, the results for the logistic map again closely align with the theoretical upper bounds $\mathcal{C}_{max}$. When increasing the embedding dimension, the estimated values approach the theoretical $\mathcal{C}_{max}$ curve more and more closely\textcolor{red}{, implying that the SCM values are (almost) completely determined by the underlying entropy values as $D$ becomes large. For $\mathcal{C}_{E}$ and $\mathcal{H}_{E}$ the obtained SCM values align particularly closely with $\mathcal{C}_{max}$ already at low embedding dimensions (Fig.~\ref{fig:CElogistic}(c, f)). We hypothesize that this is due to the fact that the definition of these characteristics takes information on both, pattern frequencies and transition frequencies into account, while the other two complexity--entropy pairs focus on only one if those two aspects.}

\textcolor{red}{More importantly, for sufficiently large $D$ the resulting complexity--entropy pairs} essentially cover exclusively the left uprising branch of the curve for the SCM \textcolor{red}{accounting for} transition frequencies (Figs.~\ref{fig:CElogistic}(b, e) and \ref{fig:CElogistic}(c, f)), which distinctively differs from the original permutation entropy based SCM. \textcolor{red}{In this regard, one may note that this branch implies an almost linear dependency between $\mathcal{C}_W$ and $\mathcal{H}_{W}$ (Fig.~\ref{fig:CElogistic}(e)), respectively, $\mathcal{C}_{E}$ and $\mathcal{H}_{E}$ (Fig.~\ref{fig:CElogistic}(f)), which especially applies for small entropy values within an increasingly wide range as $D$ increases. This linear increase of the maximum complexity versus entropy curve for sufficiently large embedding dimension can be expected from theoretical considerations~\cite{SMCbook2010}.}

In our opinion, the most interesting result is, however, that we do not find any \textcolor{red}{complexity--entropy pairs} in the lower right part of the plane, which for the classical SCM based on permutation entropy would present the challenging region of possible ambiguity between deterministic-chaotic and stochastic dynamics. {\color{red}By contrast, we find many values in the lower \emph{left} corner of the plane showing small values of both, $\mathcal{C}_{W}$ and $\mathcal{C}_{E}$ (Fig.~\ref{fig:CElogistic}(b, c, e, f)). This region of the complexity--entropy plane may be indicative of regular (periodic) orbits with low complexity. As already emphasized above, along with the choice of embedding parameters, the length of time series and the finite computational precision play important roles in the practical estimation process, possibly mimicking regular behavior in case of actually chaotic dynamics, which could be a reason for the unexpected population of the lower left part of the complexity--entropy planes. The mentioned numerical effects will be further discussed in Sec.~\ref{sec:transi}. Starting out of both, the lower left and right corners, the transition frequency based SCMs $\mathcal{C}_{W}$ and $\mathcal{C}_{E}$ show consistent trends along with the degree of chaoticity, as will also be detailed in Sec.~\ref{sec:transi} below. Taken together,} we suggest that the deterministic transition behavior among patterns allows an improved distinction from random noise \textcolor{red}{as compared to pattern frequency based approaches}.

\begin{figure*}
        \centering
        \includegraphics[width=2\columnwidth]{CompEntropyCNormalized_Logistic.pdf}
\caption{\textcolor{red}{(Color online) Normalized complexity -- entropy planes of the three SCMs for the logistic map: (a) $\mathcal{C}_O$, (b) $\mathcal{C}_W$, and (c) $\mathcal{C}_E$. In (c), normalized complexity values larger than 1 indicate numerical artifacts when the control parameter $r$ approaches a bifurcation point, resulting in long transients. The respective embedding dimensions are indicated by different colors (see legend).  \label{fig:CHlogCmax}}}
\end{figure*}

{\color{red}The behavior of the different SCMs relative to their respective lower and upper bounds can be further highlighted by considering normalized quantities. Specifically, for a given $\mathcal{H}$ and $D$, we define a normalized complexity $\mathcal{C}^{N} = (\mathcal{C} - \mathcal{C}_{min}) / (\mathcal{C}_{max} - \mathcal{C}_{min})$, which is theoretically bound to the interval $[0,1]$. Large values of those normalized complexities close to 1 indicate a good alignment with the maximum complexity curve. The corresponding results for the logistic map are shown in Fig.~\ref{fig:CHlogCmax}. As expected, for the traditional $\mathcal{C}_{O}$, the behavior is characterized by many points in the right half plane (large entropies, see Fig.~\ref{fig:CHlogCmax}(a)) for all considered embedding dimensions. In contrast, there are no points in the range of $\mathcal{H} >0.8$ for $\mathcal{C}_W$ and $\mathcal{C}_E$ (Fig.~\ref{fig:CHlogCmax}(b, c)). Note that in Fig.~\ref{fig:CHlogCmax}(c), some normalized complexity values larger than 1 arise at small entropies due to numerical artifacts when the control parameter $r$ is close to a bifurcation point, which results in long transients (as will be explained in Sec.~\ref{sec:transi}).}

\subsubsection{Fractional Brownian motion and fractional Gaussian noise}
In order to further analyze the capability of our modified complexity--entropy planes to distinguish chaotic from stochastic dynamics, we \textcolor{red}{next} compare the results for the logistic map with those obtained for fractional Brownian motion (fBm), i.e., stochastic processes with long-range temporal correlations. Specifically, for an fBm process the long-range correlations of the process are uniquely characterized by the Hurst exponent $H$ when positively correlated (persistence) for $1/2 < H < 1$, while being suppressed (anti-persistence) for $0< H < 1/2$. $H=1/2$ corresponds to the classical Brownian motion. Similar to the previous case of the logistic map with varying control parameter $r$, we generate time series for varying $H \in (0, 1)$ with a step size of $\Delta H=0.05$ and present the resulting complexity--entropy planes in Fig.~\ref{fig:CEfBm}.

Moreover, since fBm series in the persistent regime are nonstationary, we further transform all fBm realizations into stationary time series by employing first-order difference filtering, i.e., by considering the increments $x_{i+1}- x_i$. The transformed series are commonly referred to as fractional Gaussian noise (fGn), and the associated complexity--entropy planes are shown in Fig.~\ref{fig:CEfGn}. Notably, fGn retains the long-range correlations and Gaussian probability density function (PDF) of the underlying fBm process. For a discussion on practical concerns regarding time delay embedding of fBm and fGn processes, we refer to Ref.~\cite{Zou2015}.

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{CompEntropy_fBm.pdf}
\caption{\small{(Color online) Same as in Fig.~\ref{fig:CElogistic}, but for fractional Brownian motion with a Hurst exponent $H \in (0, 1)$ varying with a step size of $\Delta H=0.05$. Note that the normalized entropy is generally decreasing with increasing $H$ (red arrow).}  \label{fig:CEfBm}}
\end{figure*}
